\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{microtype}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{parskip}

\geometry{margin=1.1in}
\setlength{\parskip}{4pt}

\newtheorem{proposition}{Proposition}
\newtheorem{corollary}[proposition]{Corollary}
\newtheorem{remark}{Remark}

\title{\textbf{Activation Vector Baking — Research Progress Report}\\[0.3em]
\large Two Workshop Papers: Status, Results, and Forward Plan}
\author{Kamesh R\\\texttt{dorsalpanda4680@gmail.com}}
\date{May 2026}

\begin{document}
\maketitle
\tableofcontents
\newpage

% =======================================================================
\section{Context: Two Repos, Two Papers}

The project has two repositories.
\texttt{norms\_k\_calibration\_paper/} contains \textbf{Paper 1} (submitted):
the theoretical calibration paper, with all valid experiment results consolidated here.
\texttt{activation\_vectors\_as\_weights/} contains \textbf{Paper 2} (in
preparation): the AVAW companion paper on bias fusion and ramp schedules.
Both target the ICML 2026 Workshop on Weight-Space Symmetries.

% =======================================================================
\section{Starting Point: COLM 2026 Submission}
\label{sec:colm}

The project began with a safety-focused preprint submitted to COLM 2026.
The central observation: for input-independent steering vectors, runtime
addition is equivalent to a permanent bias modification:
\[
  h^{(\ell)} \leftarrow h^{(\ell)} + \lambda v^{(\ell)}
  \;\Longleftrightarrow\;
  b'^{(\ell)} = b^{(\ell)} + \lambda v^{(\ell)}.
\]
Vectors were extracted from loophole-targeted contrastive pairs
($\mathcal{D}_{\text{refuse}}$ vs.\ $\mathcal{D}_{\text{loophole}}$, 300+
pairs across cybercrime/fraud/hate speech/weapons) and baked into
\texttt{down\_proj} biases via a linear ramp over 30--80\% of layers with
grid-searched $\lambda_{\min}, \lambda_{\max}$.

\paragraph{Results (Qwen2.5-3B).}

\begin{center}\small
\begin{tabular}{llccc}
\toprule
Model & Config & HarmBench $\uparrow$ & Gibberish $\downarrow$ & GSM8K \\
\midrule
\multirow{3}{*}{Qwen-Instruct}
  & Baseline & 92.5\% & 0.5\% & 86.0\% \\
  & $+$High ($\lambda_{\max}{=}4.5$) & \textbf{98.5\%} & \textbf{0.0\%} & 60.0\% \\
  & $-$High & 59.0\% & 6.0\% & 77.4\% \\
Qwen-Base & $+$High & 49.0\% & 4.0\% & 70.7\% \\
\bottomrule
\end{tabular}
\end{center}

98.5\% HarmBench refusal with zero gibberish, competitive with Llama Guard 2.
Primary limitation: $-26$pp GSM8K at the High configuration.
Three questions left open: what $\lambda$ transfers to other models, why
30--80\%, and why a linear ramp.
Paper 1 resolves the first question analytically; Paper 2 resolves the second
and third.

% =======================================================================
\section{Paper 1: Norm-Calibrated Activation Steering}
\label{sec:paper1}

\subsection{Derivation}

\begin{proposition}[Rank-1 Perturbation Equivalence]
\label{prop:rank1}
Let $\hat{x} \sim \mathcal{U}(\mathcal{S}^{d-1})$.
A rank-1 weight perturbation $\Delta W = \alpha\,u v^\top$ produces expected
output shift $\alpha\sqrt{2/\pi d}$.
For residual-stream addition $h \mapsto h + K\hat{c}$ propagated through $W$:
$\mathbb{E}_{\hat{c}}[\|WK\hat{c}\|] = K\|W\|_F/\sqrt{d}$.
Setting these equal at $\alpha = \bar{\mu}_\ell$:
\[
  \boxed{K_\ell = \frac{\bar{\mu}_\ell}{\sqrt{d}}}, \qquad
  \bar{\mu}_\ell = \mathbb{E}_{x\sim\mathcal{D}_{\text{calib}}}\bigl[\|h^{(\ell)}(x)\|_2\bigr].
\]
\end{proposition}

\begin{proposition}[Spectral Integral — Pre-Norm Transformers]
\label{prop:integral}
In a pre-norm transformer,
$\bar{\mu}_\ell \approx \bar{\mu}_0 + \sum_{\ell'<\ell} c_{\ell'}\,\sigma_1(W_{\mathrm{dn},\ell'})$,
making $K_\ell$ a running integral of MLP spectral norms and predicting
$r(K_\ell,\,\sigma_1(W_{\mathrm{up},\ell})) > 0$ from local spectral correlations.
\end{proposition}

\begin{remark}[Gemma~2 Post-Norm Regime]
Gemma~2's post-sublayer RMSNorm clips every increment to
$\|\gamma^\mathrm{post}_\ell\|_\mathrm{eff}\sqrt{d}$, so
$\bar{\mu}_\ell \approx \bar{\mu}_0 + \sqrt{d}\sum_{\ell'<\ell}\|\gamma^\mathrm{post}_{\ell'}\|_\mathrm{eff}$.
The formula $K_\ell = \bar{\mu}_\ell/\sqrt{d}$ requires no modification;
what changes is the physical quantity $\bar{\mu}_\ell$ integrates — weight
spectra in pre-norm, accumulated normalization scales in dual-norm.
\end{remark}

$\bar{\mu}_\ell$ is estimated from 50 calibration prompts in one forward pass.
No hyperparameter search.

\subsection{Experiments and Results}
\label{sec:p1results}

Four instruction-tuned models (Llama~3.1-8B, Qwen~2.5-7B, Mistral-7B,
Gemma~2-9B), five behavioral axes (refusal calibration, formality, verbosity
control, uncertainty expression, sycophancy suppression), 45 contrastive pairs
per axis, 80/20 split, seed 42.

\paragraph{E1 — Norm profiling.}
\begin{center}\small
\begin{tabular}{lcc}
\toprule
Model & $\bar{K}$ (mean over layers) & $\overline{\|h^{(\ell)}\|_2}$ \\
\midrule
Llama 3.1 8B & 0.249 & 15.9 \\
Mistral 7B   & 0.151 & 9.7 \\
Qwen 2.5 7B  & 1.889 & 120.8 \\
Gemma 2 9B   & 6.488 & 388.2 \\
\bottomrule
\end{tabular}
\end{center}
Llama norms grow $84\times$ from layer 0 to 31; Gemma reaches 1514 at layer 41.
A fixed global $K$ is structurally wrong across and within models.

\paragraph{E2 — $K$-calibration validation.}
\begin{center}\small
\begin{tabular}{lccl}
\toprule
Model & $r$ & $p$ & Proxy \\
\midrule
Llama 3.1 8B & 0.769 & $2.7\!\times\!10^{-7}$ & $\sigma_1(W_\mathrm{up})$ \\
Qwen 2.5 7B  & 0.711 & $2.2\!\times\!10^{-5}$ & $\sigma_1(W_\mathrm{up})$ \\
Mistral 7B   & 0.715 & $4.2\!\times\!10^{-6}$ & $\sigma_1(W_\mathrm{up})$ \\
Gemma 2 9B   & $0.960$ ($\rho{=}0.9992$) & ${<}10^{-23}$ & $\sum\|\gamma^\mathrm{post}\|$ \\
\bottomrule
\end{tabular}
\end{center}
Pre-norm architectures: $r{=}0.71$--$0.77$ against $\sigma_1(W_\mathrm{up})$.
Gemma~2: $\sigma_1(W_\mathrm{up})$ is a poor proxy ($\rho{=}0.12$), but
cumulative post-norm scale recovers $r{=}0.960$/$\rho{=}0.9992$ — confirming
Remark 1.
The formula is architecture-agnostic; the proxy is not.

\paragraph{E3 — Weight-space alignment.}
\begin{center}\small
\begin{tabular}{lcc}
\toprule
Model & Mean ratio & Max ratio \\
\midrule
Llama 3.1 8B & 3.40$\times$ & 6.09$\times$ \\
Qwen 2.5 7B  & 4.05$\times$ & 7.80$\times$ \\
Mistral 7B   & 4.28$\times$ & 10.80$\times$ \\
Gemma 2 9B   & 3.04$\times$ & 5.90$\times$ \\
\textbf{Mean} & \textbf{3.69$\times$} & \textbf{10.80$\times$} \\
\bottomrule
\end{tabular}
\end{center}
Mean-max cosine similarity between behavioral PC1 and top-10 right singular
vectors of $W_\mathrm{dn}$, normalized by random baseline.
All 20 model--behavior cells exceed $1.0\times$.

\paragraph{E4 — Behavioral specificity control.}
Behavioral PC1 exceeds raw-activation PC1 alignment in three of four
architectures (Mistral: $+7.43\times$; Llama: $+1.26\times$; Qwen:
$+0.77\times$).
Gemma~2 shows no advantage ($-0.63\times$): post-norm normalization elevates
alignment of all structured directions, eliminating specificity.
Raw mean direction achieves $7$--$18\times$ alignment — a structural artifact
(mean residual direction is the primary axis $W_\mathrm{dn}$ writes along).

\paragraph{E5 — Permutation equivariance.}
\begin{center}\small
\begin{tabular}{lcc}
\toprule
Model & Mean cosine sim $\pm$ std & Invariant ($\tau{=}0.85$)? \\
\midrule
Llama 3.1 8B & $0.051\pm0.040$ & No \\
Qwen 2.5 7B  & $0.106\pm0.068$ & No \\
Mistral 7B   & $0.160\pm0.123$ & No \\
Gemma 2 9B   & $0.412\pm0.181$ & No \\
\bottomrule
\end{tabular}
\end{center}
50\% neuron permutation, 5 seeds.
Directions are equivariant, not invariant: they co-rotate with the weight
basis as predicted by E3.
A behavioral probe identifies a direction relative to one member of the
permutation orbit, not a functional-class invariant.

\paragraph{E6 — Partial correlation (controlling for depth).}
Both $K_\ell$ and $\sigma_1(W_\mathrm{up})$ grow monotonically with layer
depth, raising the concern that E2's correlation is a depth confound.
Partial correlation controlling for $\ell$ strengthens the result in all
models:

\begin{center}\small
\begin{tabular}{lccc}
\toprule
Model & Full $r$ & Partial $r$ (depth-controlled) & $p$ (partial) \\
\midrule
Llama 3.1 8B & 0.769 & \textbf{0.901} & $4.8\!\times\!10^{-12}$ \\
Qwen 2.5 7B  & 0.711 & \textbf{0.832} & $7.6\!\times\!10^{-8}$ \\
Mistral 7B   & 0.715 & \textbf{0.780} & $2.3\!\times\!10^{-7}$ \\
Gemma 2 9B   & 0.421 & \textbf{0.696} & $4.3\!\times\!10^{-7}$ \\
\bottomrule
\end{tabular}
\end{center}
The depth-controlled partial correlation eliminates the confound and the
relationship strengthens, confirming the spectral grounding of the formula
is not an artifact of monotone depth trends.

\paragraph{E7 — PCA component dominance.}
\begin{center}\small
\begin{tabular}{lccc}
\toprule
Model & PC1/PC2 ratio (mean) & Top-5 cumulative variance \\
\midrule
Llama 3.1 8B & 1.60 & 36.9\% \\
Qwen 2.5 7B  & 1.86 & 43.0\% \\
Gemma 2 9B   & 35.5 & 81.2\% \\
Mistral 7B   & 62.5 & 86.6\% \\
\bottomrule
\end{tabular}
\end{center}
Mistral and Gemma show extreme PC1 dominance (ratios of 62.5$\times$ and
35.5$\times$), meaning behavioral signal is essentially rank-1 for these
architectures.
Llama and Qwen are more distributed, consistent with their lower E3 alignment
ratios.

\paragraph{Appendix: Directional fidelity.}
\begin{center}\small
\begin{tabular}{lcccc}
\toprule
Model & Baseline & Raw ($K{=}1$) & PC1 ($K{=}1$) & PC1 ($K$-cal.) \\
\midrule
Llama 3.1 8B  & $0.0040$ & $0.0121$ & $0.0014$ & $\mathbf{0.0471}$ \\
Gemma 2 9B    & $0.0033$ & $0.0158$ & $0.0117$ & $0.0065^{\dagger}$ \\
Qwen 2.5 7B   & $0.0003$ & $0.0087$ & $0.0119$ & $0.0018$ \\
Mistral 7B    & $-0.0069$ & $0.0443$ & $0.0281$ & $0.0170$ \\
\midrule
Average & $0.0002$ & $0.0202$ & $0.0133$ & $0.0181$ \\
\bottomrule
\end{tabular}
\end{center}
Mean cosine shift ($\uparrow$ better), averaged over 5 behaviors.
\textbf{Sample size caveat}: $n{=}9$ pairs per axis per model; Qwen and
Mistral differences are within the $\pm$95\% CI and should be treated as
inconclusive, not null results.
Llama: K-calibration gives $+0.046$ absolute ($+3349\%$) — cleanest signal
because pre-norm residual norms span $>10\times$ within-model range.
Gemma ($\dagger$): slight loss predicted by Remark 1 (post-norm over-shoots
at $K_\ell \approx 25$).

\subsection{Paper Status}

Full \texttt{paper.tex}, figures (fig1--6 + appendix), and all result CSVs are
in \texttt{norms\_k\_calibration\_paper/}.
Downstream generation-level evaluation is in companion work (Paper 2), pending clean rerun after a bug was found in the prior LLM-judge run.

% =======================================================================
\section{Paper 2: Activation Vectors as Weights (AVAW)}
\label{sec:paper2}

Repo: \texttt{activation\_vectors\_as\_weights/}.
Full title: \emph{Activation Vectors as Weights: Persistent Nudges via Bias
Fusion for Library-Free Behavioural Alignment}.

\subsection{Core Contribution}

Paper 2 operationalises the bias-equivalence observation from the COLM
submission and packages it as a lossless, library-free deployment format:
bake the calibrated steering vector into \texttt{down\_proj} biases once;
load the resulting checkpoint anywhere with standard HuggingFace
\texttt{from\_pretrained} — no hooks, no steering library.

\textbf{Lossless guarantee.} The bias-equivalence derivation is exact for
input-independent $v$: $h + \lambda v = Wx + b + \lambda v = Wx + b'$.
Exp 02 provides numerical confirmation ($L^\infty < 10^{-5}$ on logits across
float16 and float32); that experiment is pending the GPU rerun.

\textbf{Distinction from prior work.} Compiling Activation Steering
(arXiv:2604.12359) and Steer2Edit use input-triggered weight edits that
activate conditionally.
AVAW's bias fusion is unconditional (always on) and lossless.
This is preferable for persistent alignment goals (sycophancy suppression,
refusal reinforcement) where the nudge should apply uniformly regardless of
phrasing.

\subsection{Cosine-Bell Ramp Schedule}

The COLM linear ramp (30--80\% of layers, searched $\lambda_{\min},
\lambda_{\max}$) is replaced by a principled schedule.
AVAW compares four:

\begin{center}\small
\begin{tabular}{ll}
\toprule
Schedule & $K^{(\ell)}_{\text{effective}}$ \\
\midrule
Flat (baseline) & $K_\ell$ uniformly \\
Linear & $K_\ell \cdot (\ell - \ell_s)/(\ell_e - \ell_s)$ \\
Cosine-bell & $K_\ell \cdot \tfrac{1}{2}(1 + \cos(\pi(1-t(\ell))))$, $t \in [0,1]$ over window \\
Norm-inverse & $K_\ell \propto 1/K_{\max}(\ell)$, redistributing budget from fragile layers \\
\bottomrule
\end{tabular}
\end{center}

Before any schedule, the layer collapse threshold $K_{\max}(\ell)$ is measured
by sweeping $\alpha$ on a held-out set and detecting the perplexity inflection
point; cosine-bell amplitudes are clipped to $K_{\max}(\ell)$.

\subsection{Experiments}

A prior downstream evaluation run (LLM-as-judge, 50 prompts per
behaviour) was found to have a bug and all results were discarded.
All AVAW experiments are pending a clean rerun on the GPU server.

\begin{center}\small
\begin{tabular}{clll}
\toprule
Exp & Description & Status \\
\midrule
01 & Cross-arch CKA (4$\times$4 heatmap, 5 depth fractions) & Ready to run \\
02 & Lossless compilation verification & Ready to run \\
03 & Collapse threshold $K_{\max}(\ell)$ sweep & Planned \\
04 & Ramp schedule ablation (flat/linear/cosine-bell/norm-inverse) & Planned \\
05 & TruthfulQA MC accuracy under each schedule & Planned \\
06 & Sycophancy suppression win-rate (GPT-4o judge) & Planned \\
07 & Refusal reinforcement — jailbreak success rate & Planned \\
08 & Direction transfer across architectures via Hub adapter & Planned \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Connection to Paper 1}

Paper 1 derives and validates $K_\ell = \bar{\mu}_\ell/\sqrt{d}$ via internal
geometric metrics.
Paper 2 uses that calibration as a given, substitutes $K_\ell$ into the
bias-fusion formula, and evaluates the resulting behavioral change at the
generation level.
The division of labour is clean: Paper 1 is the spectral geometry paper;
Paper 2 is the deployment paper.

% =======================================================================
\section{Summary Table}

\begin{center}\small
\begin{tabular}{p{3cm}p{5cm}p{5cm}}
\toprule
 & \textbf{Paper 1} (submitted) & \textbf{Paper 2} (in prep) \\
\midrule
Claim & $K_\ell = \bar{\mu}_\ell/\sqrt{d}$ from rank-1 equivalence & Bias fusion is lossless; cosine-bell ramp improves behavioral control \\
Eval & Spectral correlation, alignment ratio, permutation equivariance, directional fidelity & Generation benchmarks, LLM judge, ramp ablation \\
Gemma & $\rho{=}0.9992$ via post-norm proxy; formula unchanged & TBD \\
Status & \texttt{paper.tex} + figures complete & README + code complete; results pending \\
Defers & Generation-level behavioral eval & Theoretical calibration derivation \\
\bottomrule
\end{tabular}
\end{center}

% =======================================================================
\section{Open Questions}

\begin{enumerate}[topsep=2pt]
  \item Paper 1's directional fidelity shows Qwen/Mistral differences are
        within CI ($n{=}9$ pairs per axis).
        Is this underpowered, or is the null result ($K{=}1$ accidentally
        approximates the correct scale for mid-range norm profiles) itself
        publishable?

  \item The Gemma post-norm proxy ($\sum\|\gamma^\mathrm{post}\|$) recovers
        $\rho{=}0.9992$, but accessing it requires a weight inspection step not
        needed for pre-norm models.
        Does this undermine the ``no weight access required'' framing, or is
        the calibration-from-activations-only path the right default and the
        weight-proxy a theoretical explanation?

  \item For the AVAW sycophancy judge (Exp 06): GPT-4o as judge for
        sycophancy is a known confound.
        Is there a published evaluator you would recommend, or should I use
        the SycophancyEval fine-tuned classifier?

  \item The COLM $-26$pp GSM8K drop used a grid-searched $\lambda$ on 3B.
        If Paper 2's cosine-bell at analytically-derived $K$ recovers more of
        GSM8K at equivalent HarmBench performance on 7--9B models, that is the
        clearest cross-paper result.
        Is this the framing you would prioritise?

  \item Both papers target the same workshop.
        Is that appropriate given they cite each other, or should one be
        redirected to a safety-specific venue (e.g.\ SaTML)?
\end{enumerate}

\end{document}
